Open Data, Open Source and Open Standards in chemistry - The Blue Obelisk five years on

### 2011-05-13 paper.tex
Cleaned up the Business End section

Open Source provides a unique opportunity for commercial organisations to work with the community. As opposed to traditional business models which focus on maximising revenue through monetisation of all aspects of software, working with the community allows a business to maximise profits through greater community adoption and popularity. This is an ideal solution for some commercial organisations because only they can afford the significant investment in funding and time that is required to develop a successful open source project. Regardless, there are always challenges to growing a project, and there are specific issues that complicate making a business out of open source software in the scientific industry. A goal of the Blue Obelisk is to reduce these challenges and bring commercial organisations and the community together to benefit science.

Open source products attract interest because they are easily modifiable, and they can usually be used freely under a standard open source license. With proper advertising, a company can grow a user base more quickly than with a proprietary product, which requires significantly more marketing resources. Additionally, an open source product may integrate with other products that are governed by compatible open source licensing, allowing for faster development. Adopters benefit from a quickly established community of users, that not only know how to use that software, but may also have experience developing it. These benefits allow a company to create a successful product with significantly reduced cost and effort.

Successful open projects require the dedication of a number of individuals to ensure that the product is continually developed, supported and funded. In many non-commercial cases, an author will produce an open project as a hobby, and stop work on the project before the development reaches a certain quality or community interest reaches a critical mass. This is because the hobbyist eventually needs to focus on income or he/she has a change of priorities. On the other hand, commercial organisations have the means to devote resources to a project to ensure that it achieves an appropriate level of quality and adoption. In return, that commercial organisation gains recognition from that product and is the first choice for commercial support, consulting and custom development. This is the primary goal of commercial open source, as selling licenses for an open product is very unlikely to be feasible.

In the scientific industry, it is incredibly difficult to start successful open projects. This is because the expertise needed to contribute to such a project requires a higher degree of education and the population of end users is much smaller than other industries, such as entertainment. Because of this, there is an increased barrier to starting open projects in the scientific industry. The Blue Obelisk attempts to solve this issue, by uniting those interested in open source and by helping to increase awareness for new open projects. Commercial organisations can take advantage of this goal, as open products will provide a faster return on investment, and the community benefits from open software and new job opportunities.

For instance, iChemLabs distributes several scientific software solutions under various open and proprietary licensing. The license is chosen to best ensure the success of the project. iChemLabs released the ChemDoodle Web Components library under the GPL v3 license with a liberal HTML exception. iChemLabs is dedicated to developing, funding and supporting the library so that the scientific community can immediately utilise the maturing HTML5 technologies and quickly make web and mobile interfaces for scientific content. Other open source developers can benefit from this library freely and expect continued development, while commercial organisations that use the library under an appropriate license (GPL for open projects or purchasing a proprietary license for proprietary projects) will benefit from an already developed market. Of course, no one will need to waste time rewriting functionality that is already provided with quality. Several open projects have already incorporated the ChemDoodle Web Components, such as iBabel by Chris Swain\cite{iBabel}, ChemSpotlight by Geoffrey Hutchison\cite{chemspotlight} and the RSC ChemSpider\cite{chemspider_chemdoodle}.

An example of such an interaction was the donation by eMolecules, Inc.
to Open Babel of code for the canonicalisation of molecules and
fragments (Nov 2006). eMolecules is an online vendor of chemicals that
uses Open Babel under-the-hood to manage its compound collections.

Another more recent example occurred in July 2010 when Silicos, a
Belgian company that provides services
in the area of cheminformatics, released several command line
applications based on Open Babel as well as donating code to the
project. For example, the Pharao
tool released by them is a comprehensive solution for pharmacophore
searching that provides extensive support for a variety of
pharmacophore searches. Other tools released by Silicos include Sieve for
filtering by molecular property, Stripper for removing core scaffold
structures from a molecule set, and Piramid for molecular alignment
using shape determined by the Gaussian volumes as a descriptor.

Another example of Open Source
tools originating from commercial groups is the ChemCraft tool from
Molecular Networks GmBH, that does XXX.
% Really? I've only seen commercial software from them -GRH

Rather than releasing new tools, an alternative approach is to provide an
interface to existing tools. hBar Solutions has developed an online
portal, hBar Lab, for managing and performing computational chemistry
calculations in the cloud. To do so it leverages two Blue Obelisk
projects, Jmol and Open Babel, as well as the open source quantum
mechanics package MPQC.

>>>>

Open Source provides a unique opportunity for commercial organisations to work with the
cheminformatics community. Traditional business models rely on monetisation the source
code, causing every company to repeat the development by other companies, often combined
with a free (gratis) model for people working at academic institutes, to increase adoption
and allow contributions from academics. This solution defines the IP on the software
return on investment, but has the downside of investment losses due to duplication of
software and method development, which becomes visible when proprietary companies
merge.

The Blue Obelisk community, however, takes advantage that much of the investments needed
for development are either payed by academic institutes and funding schemes, and by
volunteers investing time and effort who get full access to the source code in return.
In fact, all partners get full access up front and the Open Source licending ensures
that they will have access any time in the future. As such, it functions as a social
contract between everyone to arrange the immediate return on investment. Effectively,
this approach shares the burden of the high investment in having to develop cheminformatics
software from scratch, allowing researchers and commercial partners alike to focus
on their core business, rather than the development of prerequisites. As such, the rich
collection of Open Source cheminformatics tools provided by the Blue Obelisk
greatly reduces investment up front for new companies in the cheminformatics market.

The use of Open Standards allows everyone to select those Blue Obelisk components
they find most useful, as they can easily replace one component with another providing
the same functionality, taking advantage that they use the same standards for,
for example, data exchange. This way, licensing issues are becoming a marginal
problem, allowing companies to select a license appropriate for their business
model. This too, allows a company to create a successful product with significantly
reduced cost and effort.

At the time of writing there are many commercial companies development chemistry
solutions around Open Source cheminformatics components provided by the Blue Obelisk
community. Examples of such companies include IChemLabs, IdeaConsult, Wingu, Silicos,
GenettaSoft, hBar, and Inkspot Science. Some of these merely use components, but several
actively contribute back to the Blue Obelisk project they use, or donate new
Open Source cheminformatics projects to the community.

For example, iChemLabs released the ChemDoodle Web Components library under the GPL v3
license, based on the upcoming HTML5 Open Standard. It allows making web and mobile
interfaces for chemical content. The project is already being adopted by others,
including iBabel by Chris Swain\cite{iBabel}, ChemSpotlight by
Geoffrey Hutchison\cite{chemspotlight} and the RSC ChemSpider\cite{chemspider_chemdoodle}.

Silicos has release several Open Source utilities based on
OpenBabel, such as Pharao, a tool for pharmacophore searching,
Sieve for filtering molecular structure by molecular property,
Stripper for removing core scafold structures from a molecule
set, and Piramid for molecular alignment using shape determined
by the Gaussian volumes as a descriptor. Additionally,
contributions have been made to the OpenBabel project itself.

Other companies use Blue Obelisk components and contribute patches,
smaller and larger. For example, IXELES donated the isomorphism
code in the CDK, eMolecules donated canonicalisation code to
OpenBabel, and AstraZeneca contributed code to the CDK for
signatures. This is just a very minor selection, and the author
is encouraged to contact the individual Blue Obelisk projects
for an elaborate list.
